---
title: "HW2-EDA with R - Part1: Plotting"
author: "misken"
date: "January 22, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

## https://github.com/pramesh30/PCDA
## Please refer to the hw_2files folder to view the commits


## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.
You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
county_df <- read.csv("data/county.csv")
county_MI <- read.csv("data/county_MI.csv")
```

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. Are there any NA values? What 
differences do you notice between the structure of the two data frames.

There are some NA values in county_df data frame, but none in county_MI.
county_MI has an additional field called penninsula with UP and LP values. There are 83 records of 41 variables in county_MI and 3142 records of 40 variables in county_df.


```{r}
str (county_df)
str (county_MI)

```

The county_MI data frame contains data for Michigan only and includes an
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.

### Problem 2 - Basic plots for Income

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 

Write a short paragraph describing the distribution of median incomes.

The graph shows that it is skewed towards the right. Analysing the histogram shows 3142 observations grouped according to median income ranginG from approx 25000 to 125000. The shApe of graph shows a bell curve indicating normal distribution of data (income) with maximum number of people observed at income range between(40000 - 45000).

```{r}
ggplot(data = county_df) + geom_histogram(bins = 30, mapping = aes(x = county_df$Income))

ggplot(data = county_df) + geom_boxplot(mapping = aes(x = "Income", y = county_df$Income))

ggplot(data = county_df) + geom_density(mapping = aes(x= county_df$Income))

ggplot(data = county_df) + geom_violin(mapping = aes(x = "Income", y = county_df$Income))

```

### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US.

```{r histo_totpop}
options(scipen=999)
histo_totpop <- ggplot(data = county_df) + geom_histogram(bins = 30, mapping = aes(x = county_df$TotalPop))
histo_totpop
```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}
histo_stylechanges <- ggplot(data = county_df) + geom_histogram(bins = 30, mapping = aes(x = county_df$TotalPop) , color = '#FFCC99', fill = '#FF91A4') + ggtitle("Distribution of population for US counties") + labs(y= "Number of counties", x = "Total Population", colour = county_df$TotalPop)

histo_stylechanges

```

**Hacker Extra** - Also create a kernel density plot overlaid on a histogram of
Income. 

```{r histoden}

histoden <- ggplot(county_df, aes(x=county_df$Income)) + geom_histogram(bins = 30, aes(x=county_df$Income, y = ..density..), fill = "#FF91A4", colour = "#FFCC99") + geom_density(alpha=.2, fill="#FF6666", colour = "black") + labs( x = "Income")
histoden

```


Repeat the histogram but only for counties with less than 2.5 million people.

```{r histo_lt2.5m}

county2.5 <- county_df %>% select (County, TotalPop, Income) %>% group_by(County) %>% filter(TotalPop < 2500000)
county2.5
histo_lt2.5m <- ggplot(data = county2.5) + geom_histogram (bins = 30, mapping = aes(x = county2.5$Income)) + labs( x = "Income")
histo_lt2.5m

```


### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped box plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?

LP and UP differ across peninsulas with respect to Incomes . AS seen in LP, Its evident that more then 50% values lie in the upper quartile meaning 50% of people have their incomes greater then the median income value.Thus we can conclude the the distribution in LP according to spread of income is not normal.In other words, data is skewed.

In UP, data is less skewed making the values look equally distributed across the median value meaning there are 50% people have their incomes above median value and roughly 50% of them have their incomes below median value making the distrubution look like a normal distribution.


```{r bp_inc_by_peninsula}
box_incpen <- ggplot(data = county_MI) + geom_boxplot(aes(x = county_MI$peninsula, y = county_MI$Income)) + ggtitle("Distribution of Income Vs Peninsula") + labs(y= "Income", x = "Peninsula")

box_incpen
```

Create several more boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

```{r bp_cons_by_peninsula}

ggplot(county_MI) + geom_boxplot(aes(x = county_MI$peninsula, y = county_MI$Citizen , fill = county_MI$peninsula)) + ggtitle("Distribution of Construction Vs Peninsula") + xlab("Peninsula") + ylab("Construction")

```

Percentage range for people in construction and maintenance for LP is 8-12 (more variable range) and in UP is 10-12. There is a skewness across median value (10.5) for LP which means the data is not normally distributed in these quartiles. 25 pecent of data in 2nd quartile in LP is more variable with respect to % of people in construction versus the 25 % of data in upper quartile. The median vaule for UP is around 10.5 but its distribution is skewed in such a way that data in 2nd quartile is less variable than the upper quartile.


```{r bp_cons_by_peninsula}

ggplot(county_MI) + geom_boxplot(aes(x = county_MI$peninsula, y = county_MI$Employed , fill = county_MI$peninsula)) + ggtitle("Distribution of Employed Vs Peninsula") + xlab("Peninsula") + ylab("Employed")
```

Percentage range for people Employed in the LP is 5-50 (more variable range) and in UP is 11-15. There is a skewness across median value for LP which means the data is not normally distributed in these quartiles. 25 pecent of data in 2nd quartile in LP is more variable with respect to % of people are employed versus the 25 % of data in upper quartile. 


```{r bp_walk_by_peninsula}
ggplot(county_MI) + geom_boxplot(aes(x = county_MI$peninsula, y = county_MI$Walk , fill = county_MI$peninsula)) + ggtitle("Distribution of Walk Vs Peninsula") + xlab("Peninsula") + ylab("Walk")
```

LP has less percentage (1.75 - 3 %) of people who walk to work as compared to UP (2.5 -6 %). Median values for LP is around 2.5 and distribution in 2nd and 3rd quartile is symmetric which means data is normally distributed in these quartiles.
Data is skewed and not normally distributed for UP. Median value is slightly less then 5.


```{r bp_drive_by_peninsula}
ggplot(county_MI) + geom_boxplot(aes(x = county_MI$peninsula, y = county_MI$Drive , fill = county_MI$peninsula)) + ggtitle("Distribution of Drive Vs Peninsula") + xlab("Peninsula") + ylab("Drive")


```

Number of drivers driving alone in LP range from 79-83 with median 82. This is a normal distribution becasue, a symmetric distribution can be seen in the 2nd and 3rd quartile which means the spread of values on either side of the median value is approximately same. 

### Problem 5 - How do regions of the country compare?

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how some variable of interest to you compares by region. For 
example, you could explore how the regions compare with respect to the
distribution of MeanCommute time. Make sure your graph has a title and good axis
labels.

```{r faceted_histo}

library(scales)

faceted_histo <- ggplot(county_df) + geom_histogram(data = county_df, bins = 50, aes(x=county_df$Men))  + ggtitle("Men by Peninsula") + xlab("Region") + ylab("Number of Men") + facet_wrap(~ Region, scales = "free") 

faceted_histo

```

### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Region in the US. Make sure
your graph has a title and good axis labels.

```{r bar_counties_by_region}

bar_counties_by_region <- ggplot(county_df) + geom_bar(aes(x=county_df$Region)) + labs(y= "Number of Counties", x = "Region") + ggtitle("Number of Counties VS Region") 


bar_counties_by_region

```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}

xy_inc_commute <- ggplot(data = county_MI) + geom_point(mapping = aes(x = county_MI$Income, y = county_MI$MeanCommute)) + labs(y= "Mean Commute", x = "Income")

xy_inc_commute

```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula.

```{r xy_inc_commute_peninsula}

xy_inc_commute_peninsula <- ggplot(data = county_MI, aes(x= county_MI$Income, y = county_MI$MeanCommute, color = peninsula)) + geom_point() + labs(y= "Mean Commute", x = "Income")

xy_inc_commute_peninsula

```

Your thoughts on the relationship between these variables?

The relationship between MeanCommute and Income appears to be positive, as we move from left to right, the value of income increases with respect to the values of MeanCommute. That indicates the general look of the data line is an uphill pattern

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

```{r xy_faceted_peninsula}

ggplot(data = county_MI, aes(y = county_MI$Income, x = county_MI$MeanCommute, color = peninsula)) + geom_point() + labs(x= "Mean Commute", y = "Income") + facet_wrap(~peninsula) + ggtitle("Distirbution of Mean Commute Vs Income")
```

The relationship between MeanCommute and Income appears to be positive, as we move from left to right, the value of income increases with respect to the values of MeanCommute. That indicates the general look of the data line is an uphill pattern


```{r xy_faceted_peninsula}

ggplot(data = county_MI, aes(y = county_MI$Income, x = county_MI$Women, color = peninsula)) + geom_point() + labs(x = "Women", y = "Income") + facet_wrap(~peninsula) + ggtitle("Distirbution of Women Vs Income")
```
The relationship between Women and Income appears to be positively correlated. It appears the same for both the peninsulas. The data shows that the linear correlation could be undefined.




```{r xy_faceted_peninsula}
ggplot(data = county_MI, aes(y = county_MI$Service, x = county_MI$Office, color = peninsula)) + geom_point() + labs(x = "Office", y = "Service") + facet_wrap(~peninsula) + ggtitle("Distirbution of Office Vs Service")
```
It appears that the distribution of the point for LP is more scattered between 20-25, indicating there is a low but positive coorelation, whereas, the data for UP is more random.


```{r xy_faceted_peninsula}

ggplot(data = county_MI, aes(x= county_MI$Drive, y = county_MI$Walk, color = peninsula)) + geom_point() + labs(y= "Walk", x = "Drive") + facet_wrap(~peninsula) + ggtitle("Distirbution of Walk Vs Drive")

```
The relationship between Walk and Drive appears to be negative and has low correlation, as we move from left to right, the value of walk decreases with respect to the values of drive That indicates the general look of the data line is an downhill pattern


Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}
ggplot(county_df) + geom_point(aes(x=county_df$Income, y=county_df$TotalPop)) + ggtitle("Total Population By Income") + xlab("Income") + ylab("Total Population")

```

```{r xy_pop_inc}

ggplot(county_df, aes(x=county_df$Income, y=county_df$TotalPop))+ xlab("Total Income") +ylab("Population") + geom_point(shape=1) + ggtitle("Total Population by Income for entire US")+scale_y_continuous(labels = scales::comma)+scale_x_continuous(labels = scales::comma) + coord_trans(y = "log10")

```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.

### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.